Measuring Directionality in Double-Beta Decay and Neutrino Interactions with Kiloton-Scale Scintillation Detectors

### 2014-04-21 RefereeResponse/RefereeResponseTwo_v0.tex

\documentclass[11pt]{article}

\usepackage[usenames,dvips]{pstcol}

\usepackage{color}

\usepackage{amssymb}

\usepackage{amsmath}

\usepackage{graphicx,color}

%

%\input{symbols.tex}

%

\newcounter{question}

\newcommand{\noteq}{{\vskip0.1in\bf \large \color{red} [Q:{\color{black}\addtocounter{question}{1}\arabic{question}}]~}}

\newcommand{\noteQ}{{\bf \LARGE \color{blue} [Q: {\color{black}\addtocounter{question}{1}\arabic{question}}]~}}

\newcommand{\notea}{{\bf \large \color{red} [A]~}}

\newcommand{\noteA}{{\bf \LARGE \color{blue} [A]~}}

\newcommand{\noteqq}[1]{{\bf \large \color{red} [Q: {\color{black}#1}]~}}

\newcommand{\noteaa}[1]{{\bf \large \color{red} [A: {\color{black}#1}]~}}

%

\textwidth 6.5in

\textheight 9.4in

\topmargin -0.8in

\oddsidemargin 0.0in

\evensidemargin 0.0in

\parindent 0.0in

\def\baselinestretch{1.0}

\def \Russ {Translated from Russian. Fixed. Thanks.}

\def \F    {Fixed.}

\def \FT   {Fixed. Thanks.}

\def \D    {Done.}

\def \AP   {Basically Stylistic- Author's Prerogative.}

\begin{document}

\begin{flushright}

\today

\end{flushright}

%----------------------------------------------------------------

Dear JINST Editors,\\

\parindent 1.0in

We thank the referees again for their very helpful and knowledgeable
comments. We have made all the minor changes that were suggested by referee 1 and referee 2.

The remaining point to discuss is the title of the paper, \textit{Measuring directionality in double-beta decay and neutrino interactions with kiloton-scale scintillation detectors}.  We feel that the title indicates that we will be developing a method to measure direction and  the title guides the reader to what type of event we will be proposing to reconstruct, events with $\sim$MeV energies. It is important to note that reconstruction of $\sim$GeV events in scintillating detectors has been tackled by other groups and we reference their work in this publication.

Referee 2 says ``I believe the title is sending a different message to the reader, indicating the measurement of (low energy) particle directionality as a powerful tool to improve the performance of experiments on neutrino interactions or double-beta decay."  In order to keep the title, Referee 2 requests that we address in detail the kinematics of the interactions, the Cherenkov threshold, and the spatial distribution of the decays.  We have addressed part of this request by including the number of electrons above the Cherenkov threshold and above 1.4~MeV in $^{116}$Cd neutrinoless double beta decay events. However, to address these issues further we would need to do a significant amount of work on the reconstruction algorithm. We feel goes beyond the proof of principle paper we were intending. If the editors feel strongly that we should change the title, we can, but we believe it is the correct title for the paper.

Below we list the referee requests and changes in detail.\\ \\

\parindent 3.0in

    Sincerely,\\

\parindent 4.0in

		         Lindley Winslow, for the Authors

\parindent 0.0in

%\begin{verbatim}

%\end{verbatim}

{\bf Referee 1: }

\noteq [Referee 1 1) Organization] ``Please consider organizing the paper along these items/subsections be-
low (or others equivalent), in order to avoid a pure sequential enumeration
of (disconnected) sections''

\notea{\bf We have changed the organization.}\\

\noteq [Referee 1: 2) Figures] ``In general, the figures are not really easy to find''

\notea{\bf  \F~We have worked on the figure placement}\\

\noteq [Referee 1: 3) ] ``it is very difficult for an outsider to understand why the authors are so attached to cos(theta)
in Fig.7.''

\notea{\bf We have removed Fig. 7. We were not attached to the figure.}\\

{\bf Referee 1: Specific Comments}

\noteq [Referee 1:  Abstract and Conclusions] ``Develop a technique probably too strong: ”we propose a technique” I would say. Developing it will take some more time, to my understanding...''

\notea{\bf \FT}\\

\noteq [Referee 1:  Abstract and Conclusions] ``Generally, I suggest not to assume that everybody understands what TTS (or even what transit-time spread) is, specially outside a context. In the figures, one could add *photo-detector* transit-time spread, for instance.''

\notea{\bf \FT}\\

\noteq [Referee 1:  Abstract and Conclusions] ``Fig. 3 EPS$\rightarrow$PEs''

\notea{\bf \FT}\\

\noteq [Referee 1:  Page 7, Top] ``We note that due to scattering the final... $\rightarrow$one could
add Coulomb scattering since the word scattering was used before in other
context.'' and "A bit later: has a distance from... $\rightarrow$travels a distance from''

\notea{\bf \FT}\\

\noteq [Referee 1:  Page 7, Bottom] ``after the TTS resolution has been applied $\rightarrow$probably
rephrase.''

\notea{\bf \D}\\

\noteq [Referee 1:  section 4:] ``truth information $\rightarrow$’MC truth’ information.''

\notea{\bf \FT}\\

\noteq [Referee 1:  section 4:] ``For the default simulation case... $\rightarrow$I am fine, but it is not easy to re-
member all details and a table may help.''

\notea{\bf We have added a table.}\\

\noteq [Referee 1:  Fig. 5(Left)]  ``Some strange points appear in between the p x(y,z)/p lines...''

\notea{\bf  \FT}\\

\noteq [Referee 1:  Fig. 5(Top)]  ``Fig. 5(Top) $\rightarrow$Fig. 5(Left).''

\notea{\bf \FT}\\

\noteq [Referee 1:  Conclusions]  ``We have developed a technique $\rightarrow$not quite''

\notea{\bf Changed to $\rightarrow$ We have proposed..}\\

\noteq [Referee 1:  Conclusions]  ``370nm $\rightarrow$for pseudocumene''

\notea{\bf Added for a typical scintillator.}\\

\noteq [Referee 1:  Conclusions]  ``All light shorter $\rightarrow$with wavelength shorter''

\notea{\bf \D}\\

\noteq [Referee 1:  Conclusions]  ``as a part $\rightarrow$as part?''

\notea{\bf \D}\\

\noteq [Referee 1:  Conclusions]  ``TTS $\rightarrow$Do not use acronym here, for sure, at least not only.''

\notea{\bf \D}\\

\noteq [Referee 1:  Conclusions]  ``Similarly the mean cos(theta)... please see my comment above on this re-
gard.''

\notea{\bf \F~We have removed this line as it does not add any information.}\\

\noteq [Referee 1:  Conclusions]  ``We plan to continue work on the topic $\rightarrow$please address what you plan to do
very shortly and what future steps are.''

\notea{\bf \F~We have added a sentence about our plans.}\\

\noteq [Referee 1:  Conclusions]  ``wherever$\rightarrow$whenever?'

\notea{\bf \D}\\

{\bf Referee 2: }

\noteq [Referee 2:  Physics Case]  ``The reference to the Gotthard experiment (in which the identification of the two electron tracks
was based on a different principle) is probably not straightforward and
should be made more transparent.''

\notea{\bf \F~We are now more explicit that this is a different technology.}\\

\noteq [Referee 2:  Physics Case]  ``5 MeV, head to head single electrons: the authors have added 2 more
examples at lower energies. These are more appropriate for double-beta
decay and provide useful information.''

\notea{\bf  We are glad that they are useful.}\\

\noteq [Referee 2:  Physics Case]  ``Cherenkov thresholds: a mention to the inefficiency in the identification
of both electron tracks would be worth."

\notea{\bf \D~We have calculated the efficiency for $^{116}$Cd and put that with the description of the first figure. A more detailed study is part of our future work.}\\

\noteq [Referee 2:  Page 5, line 3 from bottom] `` ``greater than 20 m'' ...The conclusion is
reasonable for an event originating in the center of the detector. The
situation could be substantially different for (a large fraction of) events
originating far from the center (close to the balloon surface) or larger
scale detectors.''

\notea{\bf \D~We now note that these effects will become important with uniformly distributed and lower energy events.}\\

\noteq [Referee 2:  Caption to Figure 3.]  ``I understand this is probably not very relevant
but I assume that the 1000 electrons generation is carried out in the
same conditions described for Figure 1. If this is the case maybe it
would be worth to specifiy it.''

\notea{\bf \D~We now specify that these are single electrons moving in the x-direction.}\\

\noteq [Referee 2:  Page 5, line 2.] ``What is the main effect of incomplete coverage? Simply
a lowering of the measured number of photons?"

\notea{\bf \D~We now specify that a uniform decrease in coverage would simply lower the measured number of photons.}\\

\noteq [Referee 2: Page 8, line 4 from bottom] ``even at 1.4 MeV” looks a quite strong
statement taken into account that this is still a favorable condition in
double-beta decay (one of the two electrons has always an energy in
the interval 0-Q/2.)''

\notea{\bf \D~We now note that the photon statistics will continue to drop as we head to lower energies.}\\

\end{document}

### 2014-04-21 RefereeResponse/RefereeResponse_v1.tex

%This is Janet's for the NSF.
\documentclass[11pt]{article}
\usepackage{graphicx,color}
\usepackage{rotating}
\usepackage{epsfig,graphics,rotate,color}
\usepackage{wrapfig}
\topmargin 0pt
\advance \topmargin by -\headheight
\advance \topmargin by -\headsep
\textheight 8.9in     
\oddsidemargin 0pt
\evensidemargin \oddsidemargin
\marginparwidth 0.5in
\textwidth 6.5in

%This is mine for changing the font.
\usepackage[T1]{fontenc}
\usepackage{times}
 \usepackage{hyperref}

%This is for NSF style
\bibliographystyle{utphys}

\begin{document} 
%Do this myself
%\begin{center}
%\Large{Next-Generation Liquid-Scintillator-Based Detectors: \\ Quantums Dots and Picosecond Timing}\\[0.25cm]
%\large{Lindley Winslow,  \today}\\[0.25cm]
%\end{center}
\noindent
Dear JINST Editors, \\

In the following draft we have attempted to address the excellent comments of the referees and below we provide some additional information. In some areas we would like to limit the scope of the work proposed by the referees. We envisioned this paper as a proof-of-principle. The reconstruction of the direction of low energy ($\sim$1 MeV) electrons in a liquid scintillator detector has never been shown before. In this paper, we wanted to show the following: the scintillation/Cherenkov separation was possible, the parameters of a detector that were important to this separation, and finally that Cherenkov reconstruction techniques would work in idealized situations. We are now working on at least two follow-up papers. The first is a detailed study of  reconstruction in more realistic scenarios. The second is a $0\nu\beta\beta$  experiment proposal where the detailed background reduction factors and sensitivity to new physics are presented, SuperNEMO has a similar paper but nothing similar exists for scintillator detectors since a directional signal has not been considered for them in the past.  Finally, we will be having a turnover of personnel so this paper provides an excellent summary of their work.\\

\noindent
The following is a more detailed response to the referees' comments.\\

\noindent
Thank you,\\ 
Lindley Winslow for the authors
\\
\\
\noindent
{\bf Response to Referee 1}\\
Page 1: \\
Liquid scintillators detectors quote their resolution in terms of  $\sim$5\%$/\sqrt{E(MeV)}$. We have updated the text to make that more clear. \\ \\
Rewrote introduction to include NEXT/SuperNEMO and to be more clear about the energy resolution issues and our approach to this paper. \\ \\
Defined PPO earlier.\\

\noindent
Page 3: \\
Added some example values for the equations.\\ \\
There is one referee comment here that we did not know how to interpret, {\it This case is shown in Fig.1.
->Indicating the direction of the electrons should not harm.}\\ \\
Changed the wording about the Q-value and the background considerations. \\ \\

\noindent
Page 4: \\
Also added here some reference values for the cutoff wavelength etc. as requested. \\ \\
We have added a more detailed explanation on why optical scattering can be neglected and a detailed reference. In Fig.~\ref{scatt} below we also show the simulation results with and without optical scattering for reference. In the simulation without scattering a ``pure" attenuation length is used so photons are simply absorbed and disappear from the simulation, therefore when scattering is turned on more photons reach the sphere. The timing distribution gains a small tail at longer times as is expected.\\ \\
We fixed the wording for the explanation of the QE values.\\ \\
We added some text explaining why 5~MeV is good for evaluating the technique.\\ \\
We added the distance between the start and stop points. As can be seen in Fig.~\ref{truthscatt}, the scattering angles are small until the electron falls below Cherenkov threshold, therefore the light emitted encodes the emission direction of the primary electron.\\ \\

\begin{figure}
        \begin{center}
        \includegraphics[scale=0.40]{Cerenkov_PEs_scattering_comparison.pdf} 
         \caption{The arrival time of Cherenkov light with (red) and without scattering (black).\label{scatt}}
        \end{center}
\end{figure}

\noindent
Page 5: \\
We rephrased the discussion of the energy transfer between solvent and solute.\\ \\
We updated the citations for the LAPPDs to more recent ones.\\ \\

\noindent
Page 6: \\
We have added a reference to the Geant4 standard electron physics list.  The main request of the referee is to see Fig. 3 as a function of the electron path. We are not setup to easily extract that information. Instead we point them to the data in Fig.~\ref{truthscatt} and we note again that the electron scattering is small during the time the electron is above Cherenkov threshold. In addition, we have remade Fig. 3 with all three energies and the Cherenkov structure is clear down to 1.4~MeV.

\begin{figure}
        \begin{center}
        \includegraphics[scale=0.50]{1p4_MeV_Direction_to_Stop.pdf} 
        \includegraphics[scale=0.50]{1p4_MeV_Direction_to_c_threshold_point.pdf} 
        \includegraphics[scale=0.50]{1p4MeV_direction_of_last_step_h.pdf} 
         \caption[]{Truth scattering data.\label{truthscatt}}
        \end{center}
\end{figure}

\noindent
Page 7-8: \\
We would like to leave the sections on scintillator in for completeness.\\ \\

\noindent
Page 9: \\
Added reconstruction plots for all energies and now note that the p$_x$ distribution is the $cos\theta$ distribution. \\ \\
The referee correctly notes that the vertex resolution is a sizable effect, so our use of the truth timing information is a non-negligible simplification. The effect is 0.15~s for the 3cm resolution. Eventually, we should simultaneously reconstruct the direction and vertex. This also addresses the issue of direction reconstruction in other sections of the detector. {\bf Should I add more here about our logic in neglecting this effect for this paper.  Since we are putting in the TTS smearing by hand,  we could simply double the transit time spread smearing we are using. We were interested in whether timing on the order of 0.1~ns, compared to the standard 1~ns, would allow us to reconstruct direction.}\\ \\
We now show the $cos\theta$ distribution for all energies. Fig.~7  is nice as you can see the performance of the reconstruction and extrapolate to higher or lower energies. For reference,  the Cherenkov threshold for  n(370nm)=1.465892 is 0.1879~MeV. This a logical choice of wavelength for defining the Cherenkov threshold since light produced below 370 nm will be absorbed before reaching the edge of the sphere since once again this is the absorption cutoff of the scintillator. \\ \\

\noindent
Conclusions:\\
We rewrote the conclusion to emphasize more that this is the first attempt to reconstruct direction in a liquid scintillator. \\ \\ \\ \\

\noindent
{\bf Response to Referee 2}\\

\begin{itemize}
\item The referee asked for more discussion on the possible sensitivity improvement. This is what we would like to work on next. It is obvious that it will but the magnitude requires a detailed sensitivity analysis and a better understanding of the final reconstruction performance. This is why we wanted to show it was possible first and then work on these more detailed and time intensive analyses. We have added some more discussion on the experience of other experiments as was requested by referee 1.
\item We agree 5~MeV is not interesting for $0\nu\beta\beta$, however it is interesting to neutrino-electron scattering interactions (also referred to in the title) and this allowed us to verify our simulation and reconstruction and then push to lower energies where lower photon statistics make the verification more difficult. In the reconstruction section we do present the results of lower energy.
\item  We agree that position dependence is going to be important. Our next step is to simultaneously reconstruct position and direction. A preliminary attempt  using all the photons to do the position reconstruction at 5~MeV showed no bias and a decrease in the resolution to 8.5~cm. The position reconstruction uncertainty is equivalent to a larger TTS smearing so our direction reconstruction results will degrade or improve depending on the vertex reconstruction. This is the content of a more detailed paper to follow.
\item The statistical fluctuations are a good point. We added discussion on this to accompany what was Fig.~7, the average $cos\theta$ plot.  {\bf I still need to address this, any suggestions?}
\end{itemize}

Most of the more detailed requests were addressed in the response to referee 1.  These are the ones unique to referee 2:

\begin{itemize}
\item Added reference for matter-antimatter asymmetry.
\item Added definition of x.
\item Clarified wording of a ``probable case", now dependent on model. We also added the reference to Boehm and Vogel.
\item Added discussion of how scattering at the sphere would affect the timing discussion.
\end{itemize}

{\bf \it This completes the comments of the referees.}

%Here is the bibliography
%\bibliography{QuantumDotWhitepaper} 
\end{document}

### 2014-04-21 RefereeResponse/RefereeResponse_v3.tex

like the obvious result one wants to see. .... Why not to disclose

this openly to allow the reader to evaluate the results in detail?''

\notea{\bf  This is our fault- the desired plot was (is) there

(now Figure 5), but was labeled as $p_x/|\vec{p}|$ rather than the obvious

$\cos\theta$ that it is. We have made this clear on the figure and in the

text. We have also added the distribution for the 2.1~MeV point. Thanks.}\\

@@

\noteq [Referee 1: Page 1] ``Please specify FWHM or sigma and give a

reference for both cases.''

\notea{\bf  \D  ~We have added references to KamLAND and Borexino.}\\

@@

\notea{\bf  \FT}\\

\noteq [Referee 1: Page 2] ``..the authors should..check and refer to

the work of NEXT and Super-Nemo collaborations:''

\notea{\bf  \D We have added a discussion of previous work on directionality.}\\

@@

\notea{\bf \D}\\

\noteq [Referee 1: Page 3]  ``Indicating the direction of the

electrons should not harm''

\notea{\bf  We were not sure what the referee was requesting. If this is a request for figure 1, we have tried adding the arrows and it becomes too busy. We think the caption description works well.}\\

@@

\noteq [Referee 1: Page 3] ``None of the best limits or claims in bb0

uses an isotope that satisfies this condition..''

\notea{\bf \FT We have rewritten this section to address this comment.}\\

@@

\noteq [Referee 1: Page 4] ``could you perhaps include the

attenuation @400nm and n@400..?''

\notea{\bf \D We have added the attenuation length at 400nm. We have also added the n at other wavelengths in other locations. Unfortunately we cannot cite one figure in the references as the index of refraction is taken from the KamLAND simulation which is based on several measurements that are in the cited references.}\\

@@

to benchmark the simulation with electrons having an energy doubling

the Qbb of the most common isotopes..''

\notea{\bf We have clarified in several places

that the 5 MeV case is for neutrino-electron scattering experiments. For $0\nu\beta\beta$ we

@@

(equivalently, energy or distance, including several histograms for

each case. It would be nice to see at which energy scale or length the

electron isotropizes and the Cherenkov cone smears out.''

\notea{\bf  In the updated figure 4 (angular PE hit distribution after time cut), we added the histograms for 1.4~MeV and 2.1~MeV. This shows that also for 2.1 and 1.4~MeV the PE hits are not isotropized, including the full history of the electron tracks. We have also added the path lengths and total distance and time from the vertex to the text. We also present in Fig.~\ref{truthscatt} of this report  more raw data for the referees which supports the statement that the scattering angle is small while the majority of the Cherenkov light is produced.}\\

\begin{figure}
        \begin{center}
        \includegraphics[scale=0.50]{Figure1_1.pdf} 
        \includegraphics[scale=0.50]{Figure1_2.pdf} 
         \caption[]{"For electrons with and initial energy of 1.4~MeV, we show (Top) the angle between the initial electron direction and the direction at 0.7 MeV (=half of the initial electron energy) and (Bottom) the angle between the initial electron direction and the direction of the last Geant4 step before the electron stops. We note that due to scattering the final direction of the electron before it stops does not correspond to the initial direction; however the scattering angle is small while the majority of Cherenkov light is produced.}
        \end{center}
\end{figure}

@@

\notea{\bf  \D}\\

\noteq [Referee 1: Page 9] ``..the vertex reconstruction has to be

obtained from all photons and the direction from the Cherenkov photons..''

\notea{\bf We agree with the referee and plan to address this in future papers. To do a credible and useful job on the position dependence will require a fairly-detailed vertex reconstruction. This is beyond the scope of this present ‘proof-of-principle’ paper. We have tried to clarify this in the text }\\

@@

\notea{\bf Yes. We included more data from the lower energy studies,

  have extracted much more detail from the simulation, and have

  extensively rewritten the manuscript. It is still at the

  `proof-of-principle level. Our next work will include a full sensitivity analysis including \a deeper understanding of the performance of the reconstruction algorithms with a more detailed simulation. This is a big job and we are working on it! }\\

@@

effect of statistical fluctuations for the low energy tails.. are not

discussed. The same holds for position and direction reconstruction

capabilities''

\notea{\bf Statistical fluctuations are addressed by the studies at lower energies and we mention explicitly in the conclusions the widening of the distributions due to decreased statistics. Threshold effect are outside of the scope of this paper. In general in these detectors, the largest loss of photon statistics is a the photocathode. The electronics for PMT readout are very efficient. }\\

@@

  unfortunately, fast compared to a post-doc tenure.}\\

\noteq [Referee 2: Page 6] ``The dependence on the position and the

corresponding transit time variations cannot in my opinion be neglected.''

\notea{\bf This part of the paper describes a simplified test simulation (where the photodetector TTS was switched off) to extract the time scales of various effects. We made changes in the text to make this clear.}\\

@@

information... The comparison with a single event (or equivalently the

distribution of the variances) and the distribution for 1 MeV

electrons would improve to understand also the limits/power of the method''

\notea{\bf  We have added to what is now figure 4 the distributions for the two lower energies. We think this is a nice illustration of the limits/power of the method.}\\

@@

\noteq [Referee 2: reconstruction] ``It is not clear if limiting the

reconstruction to Cherenkov photons improve reconstruction''

\notea{\bf Yes, we agree. In this paper we were focussing on the Cherenkov photons as a simple way to apply direction reconstruction algorithms. In the future we will work to reconstruct position and direction simultaneously from the scintillation and Cherenkov light.}\\
